---
title: "Heart failure survival"
format: html
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  error: false
---

This report uses the [Heart Failure Clinical Records](https://archive.ics.uci.edu/dataset/519/heart+failure+clinical+records) dataset to compare clinical metrics across patient groups, focusing on younger patients' survival outcomes and diabetes status.

## Connect to Data

We start by connecting to the `HEART_FAILURE` database in Snowflake, standardizing column names, and filtering to our age range of interest.

```{r}
library(tidyverse)
library(DBI)
library(dbplyr)
library(gt)
library(gtExtras)
library(odbc)

# Connect to the database
conn <-
  DBI::dbConnect(
    odbc::snowflake(),
    warehouse = "DEFAULT_WH",
    database = "HEART_FAILURE",
    schema = "PUBLIC"
  )

# Connect to table
heart_failure <- tbl(conn, "HEART_FAILURE")

# Standardize column names
heart_failure <- 
  heart_failure |> 
  rename_with(str_to_lower)
```

```{r}
heart_failure <-
  heart_failure |> 
  filter(age < 50) |> 
  select(age, diabetes, serum_sodium, serum_creatinine, sex, death_event)
```

### Age Distribution by Diabetes Status and Survival

The box plot below illustrates the distribution of ages among heart failure patients, grouped by diabetes status and survival outcome (0 = Survived, 1 = Died).

```{r}
heart_failure |> 
  mutate(
    death_event = as.character(death_event), 
    diabetes = as.character(diabetes)
  ) |> 
  ggplot(aes(y = serum_sodium, color = diabetes, x = death_event)) +
  geom_boxplot() +
  labs(
    title = "Serum Sodium Levels by Diabetes Status and Survival Outcome",
    x = "Survival Outcome (0 = Survived, 1 = Died)",
    y = "Serum Sodium (mEq/L)",
    color = "Diabetes"
  ) +
  theme(legend.position = "bottom")
```

## Compare Metrics in a Table with `{gt}`

In the table below, we compare clinical metrics across patient groups categorized by survival outcome ("Survived" or "Died") and diabetes status ("Yes" or "No"). This breakdown provides insights into how these clinical characteristics differ based on patient health conditions.

```{r}
comparison <- 
  heart_failure |> 
  group_by(death_event, diabetes) |> 
  summarize(
    median_age = median(age, na.rm = TRUE),
    median_serum_creatinine = median(serum_creatinine, na.rm = TRUE),
    median_serum_sodium = median(serum_sodium, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
comparison |> 
  mutate(
    death_event = case_when(
      death_event == 1 ~ "Died",
      death_event == 0 ~ "Survived"
    ),
    diabetes = case_when(
      diabetes == 1 ~ "Yes",
      diabetes == 0 ~ "No"
    )
  ) |> 
  arrange(desc(death_event), desc(diabetes)) |> 
  gt(rowname_col = "death_event") |> 
  fmt_number(decimals = 1) |> 
  cols_label(
    diabetes = "Diabetes Status",
    median_age = "Median Age",
    median_serum_creatinine = "Median Serum Creatinine (mg/dL)",
    median_serum_sodium = "Median Serum Sodium (mEq/L)"
  ) |> 
  tab_header(
    title = "Clinical Metrics by Survival Outcome and Diabetes Status"
  ) |> 
  data_color(
    columns = c(median_serum_creatinine, median_serum_sodium),
    palette = "Blues"
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups(groups = everything())
  )
```

